---
title: "Continuous summary tables in R"
description: >
  Build continuous summary tables in R with table_continuous(),
  including grouped descriptives, group-comparison tests, effect sizes,
  and export to console, gt, tinytable, flextable, Excel, or Word.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous summary tables in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

pkgdown_dark_gt <- function(tab) {
  tab |>
    gt::opt_css(
      css = paste(
        ".gt_table, .gt_heading, .gt_col_headings, .gt_col_heading,",
        ".gt_column_spanner_outer, .gt_column_spanner, .gt_title,",
        ".gt_subtitle, .gt_sourcenotes, .gt_sourcenote {",
        "  background-color: transparent !important;",
        "  color: currentColor !important;",
        "}",
        sep = "\n"
      )
    )
}
```

```{r setup}
library(spicy)
```

`table_continuous()` summarizes continuous variables either overall or
by a categorical grouping variable. It is designed for readable summary
tables in the console and for publication-ready outputs in rendered
documents. When `by` is supplied, it can also add group-comparison
tests, test statistics, and effect sizes.

## Basic usage

Use `select` to choose the continuous variables you want to summarize:

```{r basic}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health)
)
```

If you omit `select`, `table_continuous()` scans the data frame and
keeps numeric columns:

```{r default-select}
table_continuous(sochealth)
```

## Grouped summaries

Add `by` to summarize the same variables across categories:

```{r grouped}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education
)
```

This is the main pattern for reporting continuous variables across
groups such as education, sex, treatment arm, or survey wave.

## Add group-comparison results

Grouped tables can include inferential results directly in the same
table. The default test is Welch's test:

```{r pvalue-effect}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  p_value = TRUE,
  statistic = TRUE,
  effect_size_ci = TRUE
)
```

Use `test = "student"` for equal-variance parametric tests or
`test = "nonparametric"` for rank-based comparisons:

```{r nonparametric}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  by = education,
  test = "nonparametric",
  p_value = TRUE,
  statistic = TRUE,
  effect_size = TRUE
)
```

When you need the underlying columns for further processing, use
`output = "data.frame"`:

```{r raw-output}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score),
  by = education,
  p_value = TRUE,
  statistic = TRUE,
  effect_size = TRUE,
  output = "data.frame"
)
```

## Selecting variables

`select` supports tidyselect helpers:

```{r tidyselect}
table_continuous(
  sochealth,
  select = starts_with("life_sat"),
  by = sex
)
```

For more programmatic selection, set `regex = TRUE`:

```{r regex}
table_continuous(
  sochealth,
  select = "^life_sat",
  regex = TRUE,
  by = education,
  output = "data.frame"
)
```

Use `exclude` when you want a broad selection with one or two explicit
removals:

```{r exclude}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health, life_sat_work),
  exclude = "life_sat_work",
  by = sex
)
```

## Labels and output formats

Use `labels` to replace technical variable names with reporting labels:

```{r labels}
pkgdown_dark_gt(
  table_continuous(
    sochealth,
    select = c(bmi, wellbeing_score, life_sat_health),
    by = education,
    labels = c(
      bmi = "Body mass index",
      wellbeing_score = "Well-being score",
      life_sat_health = "Satisfaction with health"
    ),
    output = "gt"
  )
)
```

`table_continuous()` supports the same reporting-oriented outputs as
`table_categorical()`:

| `output` value | Returned object |
|:--|:--|
| `"default"` | Styled ASCII console table |
| `"tinytable"` | Formatted tinytable |
| `"gt"` | Formatted gt table |
| `"flextable"` | Formatted flextable |
| `"excel"` | Written `.xlsx` file |
| `"clipboard"` | Copied text table |
| `"word"` | Written `.docx` file |

For example, `tinytable` works well in Quarto and R Markdown documents:

```{r tinytable}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "tinytable"
)
```

## Export to Excel or Word

Use the same function to export the table directly:

```{r export, eval = FALSE}
table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "excel",
  excel_path = "table_continuous.xlsx"
)

table_continuous(
  sochealth,
  select = c(bmi, wellbeing_score, life_sat_health),
  by = education,
  output = "word",
  word_path = "table_continuous.docx"
)
```

## See also

- See `vignette("table-categorical", package = "spicy")` for grouped
  tables of categorical variables.
- See `vignette("summary-tables-reporting", package = "spicy")` for a
  cross-function reporting workflow using both summary-table helpers.
