---
title: "Categorical summary tables in R"
description: >
  Build categorical summary tables in R with table_categorical(),
  including grouped cross-tabulations, effect sizes, confidence
  intervals, and export to gt, tinytable, flextable, Excel, or Word.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Categorical summary tables in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

pkgdown_dark_gt <- function(tab) {
  tab |>
    gt::opt_css(
      css = paste(
        ".gt_table, .gt_heading, .gt_col_headings, .gt_col_heading,",
        ".gt_column_spanner_outer, .gt_column_spanner, .gt_title,",
        ".gt_subtitle, .gt_sourcenotes, .gt_sourcenote {",
        "  background-color: transparent !important;",
        "  color: currentColor !important;",
        "}",
        sep = "\n"
      )
    )
}
```

```{r setup}
library(spicy)
```

`table_categorical()` builds publication-ready categorical tables suitable for
APA-style reporting in social science and data science research. With
`by`, it produces grouped cross-tabulation tables with chi-squared
\(p\)-values, effect sizes, confidence intervals, and multi-level
headers. Without `by`, it produces one-way frequency-style tables for
the selected variables. Export to gt, tinytable, flextable, Excel, or
Word. This vignette walks through the main features.

## Basic usage

For grouped tables, provide a data frame, one or more selected
variables, and a grouping variable:

```{r basic}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education
)
```

The default output is `"default"`, which prints a styled ASCII table to
the console. Use `output = "data.frame"` to get a plain numeric
data frame suitable for further processing.

## One-way tables

Omit `by` to build a frequency-style table for the selected variables:

```{r oneway}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  output = "default"
)
```

## Output formats

`table_categorical()` supports several output formats. The table below
summarizes the options:

| Format | Description |
|---|---|
| `"default"` | Styled ASCII table in the console (default) |
| `"data.frame"` | Wide data frame, one row per modality |
| `"long"` | Long data frame, one row per modality x group |
| `"gt"` | Formatted gt table |
| `"tinytable"` | Formatted tinytable |
| `"flextable"` | Formatted flextable |
| `"excel"` | Excel file (requires `excel_path`) |
| `"clipboard"` | Copy to clipboard |
| `"word"` | Word document (requires `word_path`) |

### gt output

The `"gt"` format produces a table with APA-style borders, column
spanners, and proper alignment:

```{r gt}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = c(smoking, physical_activity, dentist_12m),
    by = education,
    output = "gt"
  )
)
```

### tinytable output

```{r tinytable}
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = sex,
  output = "tinytable"
)
```

### Data frame output

Use `output = "data.frame"` for a wide numeric data frame (one row per
modality), or `output = "long"` for a long format (one row per
modality x group):

```{r data-frame}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  output = "data.frame"
)
```

## Custom labels

By default, `table_categorical()` uses variable names as row headers. Use the
`labels` argument to provide human-readable labels:

```{r labels}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = c(smoking, physical_activity),
    by = education,
    labels = c("Smoking status", "Regular physical activity"),
    output = "gt"
  )
)
```

## Association measures and confidence intervals

By default, `table_categorical()` reports Cramer's V for nominal variables and
automatically switches to Kendall's Tau-b when both variables are
ordered factors. Override with `assoc_measure`:

```{r assoc-measure}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  assoc_measure = "lambda",
  output = "tinytable"
)
```

Add confidence intervals with `assoc_ci = TRUE`. In rendered formats
(gt, tinytable, flextable), the CI is shown inline:

```{r ci-rendered}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = c(smoking, physical_activity),
    by = education,
    assoc_ci = TRUE,
    output = "gt"
  )
)
```

In data formats (`"data.frame"`, `"long"`, `"excel"`, `"clipboard"`),
separate `CI lower` and `CI upper` columns are added:

```{r ci-data}
table_categorical(
  sochealth,
  select = smoking,
  by = education,
  assoc_ci = TRUE,
  output = "data.frame"
)
```

## Weighted tables

Pass survey weights with the `weights` argument. Use `rescale = TRUE` so
the total weighted N matches the unweighted N:

```{r weighted}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = c(smoking, physical_activity),
    by = education,
    weights = "weight",
    rescale = TRUE,
    output = "gt"
  )
)
```

## Handling missing values

By default, rows with missing values are dropped (`drop_na = TRUE`).
Set `drop_na = FALSE` to display them as a "(Missing)" category:

```{r missing}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = income_group,
    by = education,
    drop_na = FALSE,
    output = "gt"
  )
)
```

## Filtering and reordering levels

Use `levels_keep` to display only specific modalities. The order you
specify controls the display order, which is useful for placing
"(Missing)" first to highlight missingness:

```{r levels-keep}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = income_group,
    by = education,
    drop_na = FALSE,
    levels_keep = c("(Missing)", "Low", "High"),
    output = "gt"
  )
)
```

## Formatting options

Control the number of digits for percentages, p-values, and the
association measure:

```{r formatting}
pkgdown_dark_gt(
  table_categorical(
    sochealth,
    select = smoking,
    by = education,
    percent_digits = 2,
    p_digits = 4,
    v_digits = 3,
    output = "gt"
  )
)
```

## Exporting to Excel, Word, or clipboard

For Excel export, provide a file path:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "excel",
  excel_path = "my_table.xlsx"
)
```

For Word, use `output = "word"`:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity, dentist_12m),
  by = education,
  output = "word",
  word_path = "my_table.docx"
)
```

You can also copy directly to the clipboard for pasting into a
spreadsheet or a text editor:

```r
table_categorical(
  sochealth,
  select = c(smoking, physical_activity),
  by = education,
  output = "clipboard"
)
```
